\chameleon{}: Untangling Type Errors Through Interactive Visualization and Exploration

### 2022-12-12 main.tex
Change figure heads

Dynamically-typed programming languages are popular in education and the software industry. While presenting a low barrier to entry, they suffer from run-time type errors and longer-term problems in code quality and maintainability. Statically typed languages, while showing strength in these aspects, lack in learnability and ease of use. In particular, fixing type errors poses challenges to both novice users and experts. Further, compiler type error messages are presented in a static way that is biased towards the first occurence of the error in the program code. To help users resolve such type errors we introduce \chameleon{}, a type debugging tool that presents type errors to the user in an unbiased way, allowing them to explore the full context of where the errors could occur. Programmers can interactively verify the steps of reasoning against their intention. Through three studies involving real programmers, we showed that  \chameleon{} is more effective in fixing type errors than traditional text-based error messages. This difference is more significant in harder tasks. Further, programmers actively using \chameleon{}'s interactive features are shown to be more efficient in fixing type errors than passively reading the type error output.
\end{abstract}

>>>>

Dynamically typed programming languages are popular in education and the software industry. While presenting a low barrier to entry, they suffer from run-time type errors and longer-term problems in code quality and maintainability. Statically typed languages, while showing strength in these aspects, lack in learnability and ease of use. In particular, fixing type errors poses challenges to both novice users and experts. Further, compiler type error messages are presented in a static way that is biased toward the first occurrence of the error in the program code. To help users resolve such type errors we introduce \chameleon{}, a type debugging tool that presents type errors to the user in an unbiased way, allowing them to explore the full context of where the errors could occur. Programmers can interactively verify the steps of reasoning against their intention. Through three studies involving real programmers, we showed that  \chameleon{} is more effective in fixing type errors than traditional text-based error messages. This difference is more significant in harder tasks. Further, programmers actively using \chameleon{}'s interactive features are shown to be more efficient in fixing type errors than passively reading the type error output.
\end{abstract}

### 2022-12-12 sections/chameleon.tex
Change figure heads

\begin{figure}
    \centering
    \includegraphics[width=\linewidth,trim=0mm 35mm 0mm 0mm]{images/add-pair-example.pdf}
    \caption{
With traditional type error messages, one expression is blamed for causing the type error. This expression may not always match the programmers' intention. 
    }
    \label{fig:motivation-example}
\end{figure}
\subsubsection{\textbf{Traditional type errors show only limited location}}
Haack and Wells~\cite{haack_type_2004} noted that ``\textit{Identifying only one node or subtree of the program as the error location makes it difficult for programmers to understand type errors. To choose the correct place to fix a type error, the programmer must find all the other program points that participate in the error.}''  In the example of Fig \ref{fig:motivation-example}, the type errors can be fixed in multiple locations. For instance  replacing \texttt{['0'..'9']} on line 1 with \texttt{[0..9]}, or  replacing \texttt{fst x} and \texttt{snd x} on line 2 with \texttt {read (fst x)} and \texttt{read (snd  x)}. In the type error message, only the \texttt{addPair} expression on line 4 was blamed.  In this small example the whole context is visible, but the problem is worse for the programmer in large programs where the lines of code contributing to the type error are far apart in the source code.

@@

\subsubsection{\textbf{Traditional type errors give poor explanations}}
When the compiler rejects a program, the internal state of type checking is the result of complex computation. But the details of this process are hard to explain to users and usually not reported by compilers. For the typical type error shown in Fig.~\ref{fig:motivation-example}, the evidence for the type error is gathered from the previous two declarations. These have to be rediscovered by the programmer, using less rigourous methods.

@@

\section{Chameleon IDE} \label{chameleon}
\begin{figure*}
    \centering
    \includegraphics[width=\textwidth, trim=0mm 10mm 0mm 0mm]{images/atonomy.pdf}
    \caption{
        \textbf{The anatomy of \chameleon{}.}
        %\chameleon{} consists of two panes: an editor pane on the left and a
        %debugging pane on the right. 
        The editor pane (left) is similar to a
        traditional code editor. Fragments of source code may have a highlight
        color (A). Additionally, an explanation layer (B) displays in the editor
        window, if deduction steps are enabled. The debugging window contains three blocks. 
        An error statement and configuration block, containing
        An error statement (D), and, optionally,
        a list of candidate expression cards (E), a list of deduction steps (F), 
        and a control bar (G) to increment/decrement deduction step. 
        The conflicting types
        block shows two alternative types (H) inferred by the type inference
        engine. A relevant type information block to show additional information (I)
        that may help understand type errors.
    }
    \label{fig:anatomy}
\end{figure*}

@@

The \chameleon{} debugging interface provides three main features to visualize the relation between error statement and locations in code and to explore the explanation of each error location given by the type inference engine. %Further, programmers  can turn on and off features to suit their preferences and debugging needs.

@@

The type compare tool shows conflicting types in different colors, each associated with one or more error locations highlighted in a matching colour (Fig.~\ref{fig:compare}).  
%The type error must be fixed by modifying at least one of these locations. 
%Error locations are highlighted with the matching color of the conflicting types. The type compare tool is useful to quickly bisect the type error. 
If the programmers know the expression's intended type (they usually do), they will be able to eliminate half of the possible locations. 
%To make this bisecting effect more pronounced, a user can hover over one of the possible types and only show the relevant locations that contribute to that type. 
A hover interaction over one of the possible types facilitates such bisection, causing only the relevant locations that contribute to that type to be highlighted. 
%This is a convenient way to put the error \emph{under the spotlight}.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth, trim=0mm 10mm 0mm 0mm]{images/intro-candidate.pdf}
    \caption{
        \textbf{\chameleon{} with only candidate expression cards enabled.}
        Indicates the type error can
        occur in the definition of \texttt{x} or \texttt{y}.
    }
    \label{fig:expression}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=\linewidth, trim=0mm 10mm 0mm 0mm]{images/intro-deduction.pdf}
    \caption{
        \textbf{\chameleon{} with both candidate expression cards and deduction steps
        enabled.}
        \chameleon{} explains the type error in four steps. In the screenshot, the active step is step 2, where \chameleon{} shows that the expression \texttt{x} and \texttt{y} should have the same type. 
    }
    \label{fig:deduction}
\end{figure}

@@

% We propose candidate expression cards to make it clear to the user that there are multiple ways to resolve a type conflict. By contrast,  standard compiler error messages such as those of GHC arbitrarily focus the users' attention on only one candidate expression, e.g.\ ``\textit{Couldn't match expected type ‘Char’ with actual type ‘Bool’ in expression x}" where x is the candidate expression. In practice, this candidate expression often does not match programmers' intention (Fig. \ref{fig:single-candidate}).  
A candidate expression is an expression that can be inferred to have two conflicting types. 
When a type error is detected, \chameleon{} provides a list of all candidate expressions and programmers are free to choose the problem to resolve by clicking on one candidate expression card. In the example shown in Fig. \ref{fig:expression}, \texttt{x} and \texttt{y} are both candidate expressions. Fixing either type error will make both expressions well-typed.

@@

Internally, deduction steps are different ways to group the error locations into two highlighted colors. If all the highlights are placed in a list-like data structure, each increment of the step numbers moves the end of one color and the start of the other forward (Fig.~\ref{fig:step-interface} right).

@@

Nielson pointed out  that the two most important issues in designing for usability are understanding the users' tasks and the differences in users \cite{jakob_nielsen_usability_1993}. From analyzing how users use \chameleon{}, we realized that the ideal debugging interface should adapt to the specific programmer and programming task. It is common that a programmer wants the debugger to just ``show the answer", and it is also possible a programmer wants to dive deeper into the problem domain and search for the optimal solution. To accommodate the need to customize the level of information density and granularity of control, \chameleon{} provides three modes: basic, balanced, and advanced. Programmers can switch between modes by clicking on the mode switching toggles (Fig. \ref{fig:anatomy}-C). The features accessible from different modes are summarised in the following table:%shown in table~\ref{tab:chameleon-features}.

@@

Using only constraints from the MUS is sufficient to locate the type error, but to recover types from type errors we need  constraints from parts of the program that are irrelevant to the type error.  For instance, consider an ill-typed 2-tuple where two possible types can be assigned: \texttt{(Int, Int)} and \texttt{(Int, String)}. The types reconstructed from Chameleon may be \texttt{(a, Int)} and \texttt{(a, String)}. Although the recovered types are consistent, they introduce a new type variable \texttt{a} making the error message harder to understand.  To solve this issue in \chameleon{}, we first expand the MUS to a maximally satisfiable subset. We do so by removing one constraint from the MUS to get a consistent set, and then repeatedly adding constraints from the global constraint store while maintaining consistency. This process stops when no more constraints can be added. The resulting subset, while not helpful to inform error locations, will produce the most concrete conflicting types, see Fig.~\ref{fig:compare-to-original}.

@@

This has a few benefits. First, concrete types provide extra information to programmers. With a type of \texttt{(Float, Float)}, programmers may want to convey a point in 2d space. However, this information is not preserved in \texttt{(a, Float)}. Second, with concrete types, the debugging front end has more flexibility to change the representation of types. If the front end understands the type language in Haskell, it is possible to recognize and replace ground facts that are overlapping in both possible types and have the same effect. The other direction is impossible. \tf{This sentence is not very clear}

@@

% \begin{lstlisting} [
%     language=Haskell, caption={
%     A simple program that is ill-typed. This program generates two constraints from line 1 and one constraint from line 2.
% }, label={listing:ifelse}
% ]
% if a then b else c
% a = "True"
% \end{lstlisting}
\begin{listing}[!ht]
\begin{minted}{haskell}
    if a then b else c
    a = "True"
\end{minted}
\vspace{-3mm}
\caption{A simple program that is ill-typed. It generates two constraints from line 1 and one constraint from line 2. }
\label{listing:1}
\end{listing}
For instance, for the program in the listing \ref{listing:1}, \chameleon{}  generates the following constraints and labels (in brackets) $T_a = Bool$ (if condition), $T_b = T_c$ (if branches), $T_a= String$  (definition). Clearly, as $T_a$ can not unify with both \textit{Bool} and \textit{String}, this program is not well typed. \chameleon{} can construct a human-readable explanation from the MUS, for example, \texttt{a} has type \texttt{Bool} because \texttt{a} is the condition of an if statement; however, \texttt{a} has type \texttt{String} because \texttt{a} is defined as the string literal \texttt{"True"}. This low-level explanation is used in a high-level user interface described in section \ref{sub:deduction-steps} to help programmers reason about the validity of error locations.

>>>>

\begin{figure}
    \centering
    \includegraphics[width=\linewidth,trim=0mm 35mm 0mm 0mm]{images/add-pair-example.pdf}
    \caption{
    A type error displayed in Visual Studio Code\cite{microsoft_visual_nodate} and the Haskell Vscode extension\cite{haskell_haskell_nodate}.
The expression \texttt{addPair} is blamed for causing the type error. This may not match the programmers' intention. 
    }
    \label{fig:motivation-example}
\end{figure}
\subsubsection{\textbf{Traditional type errors show only limited location}}
Haack and Wells~\cite{haack_type_2004} noted that ``\textit{Identifying only one node or subtree of the program as the error location makes it difficult for programmers to understand type errors. To choose the correct place to fix a type error, the programmer must find all the other program points that participate in the error.}''  In the example of Fig \ref{fig:motivation-example}, the type errors can be fixed in multiple locations. For instance  replacing \texttt{['0'..'9']} on line 1 with \texttt{[0..9]}, or replacing \texttt{fst x} and \texttt{snd x} on line 2 with \texttt {read (fst x)} and \texttt{read (snd  x)}. In the type error message, only the \texttt{addPair} expression on line 4 was blamed.  In this small example, the whole context is visible, but the problem is worse for the programmer in large programs where the lines of code contributing to the type error are far apart in the source code.

@@

\subsubsection{\textbf{Traditional type errors give poor explanations}}
When the compiler rejects a program, the internal state of type checking is the result of complex computation. But the details of this process are hard to explain to users and are usually not reported by compilers. For the typical type error shown in Fig.~\ref{fig:motivation-example}, the evidence for the type error is gathered from the previous two declarations. These have to be rediscovered by programmers using less rigorous methods.

@@

\section{Chameleon IDE} \label{chameleon}
\begin{figure*}
    \centering
    \includegraphics[width=\textwidth, trim=0mm 10mm 0mm 0mm]{images/atonomy.pdf}
    \caption{
        \textbf{The anatomy of \chameleon{}.}
        The editor pane (left) is similar to a traditional code editor. Fragments of source code may have a highlight
        color (A). Additionally, an explanation layer (B) displays if deduction steps are enabled. The debugging pane contains three blocks. First, the error statement block contains an error statement (D), optionally, a list of candidate expression cards (E), a list of deduction steps (F), and a control bar (G) to increment/decrement deduction step. Second, the conflicting types block shows two alternative types (H). Third, the relevant type information block shows additional information (I) that may help understand type errors.
    }
    \label{fig:anatomy}
\end{figure*}

@@

The \chameleon{} debugging interface provides three main features to visualize the relation between the error statement and locations in code and to explore the explanation of each error location given by the type inference engine. %Further, programmers  can turn on and off features to suit their preferences and debugging needs.

@@

The type compare tool shows conflicting types in different colors, each associated with one or more error locations highlighted in a matching color (Fig.~\ref{fig:compare}).  
%The type error must be fixed by modifying at least one of these locations. 
%Error locations are highlighted with the matching color of the conflicting types. The type compare tool is useful to quickly bisect the type error. 
If the programmers know the expression's intended type (they usually do), they will be able to eliminate half of the possible locations. 
%To make this bisecting effect more pronounced, a user can hover over one of the possible types and only show the relevant locations that contribute to that type. 
A hover interaction over one of the possible types facilitates such bisection, causing only the relevant locations that contribute to that type to be highlighted. 
%This is a convenient way to put the error \emph{under the spotlight}.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth, trim=0mm 10mm 0mm 0mm]{images/intro-candidate.pdf}
    \caption{
        \textbf{\chameleon{} with candidate expression cards enabled.}
        Indicates the type error can occur in the definition of \texttt{x} or \texttt{y}.
    }
    \label{fig:expression}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=\linewidth, trim=0mm 10mm 0mm 0mm]{images/intro-deduction.pdf}
    \caption{
        \textbf{\chameleon{} with deduction steps enabled.}
        \chameleon{} explains the type error in four steps. In the screenshot, the active step is step 2, where \chameleon{} shows that the expression \texttt{x} and \texttt{y} should have the same type. 
    }
    \label{fig:deduction}
\end{figure}

@@

% We propose candidate expression cards to make it clear to the user that there are multiple ways to resolve a type conflict. By contrast,  standard compiler error messages such as those of GHC arbitrarily focus the users' attention on only one candidate expression, e.g.\ ``\textit{Couldn't match expected type ‘Char’ with actual type ‘Bool’ in expression x}" where x is the candidate expression. In practice, this candidate expression often does not match programmers' intention (Fig. \ref{fig:single-candidate}).  
A candidate expression is an expression that can be inferred to have two conflicting types. 
When a type error is detected, \chameleon{} provides a list of all candidate expressions, and programmers are free to choose the problem to resolve by clicking on one candidate expression card. In the example shown in Fig. \ref{fig:expression}, \texttt{x} and \texttt{y} are both candidate expressions. Fixing either type error will make both expressions well-typed.

@@

Internally, deduction steps are different ways to group the error locations into two highlighted colors. All the relevant constraints (highlights) are divided into two groups, denoted by the two colors. Each color infers a different type of the candidate expression. Each increment of the step changes the place where the two colors are divided by moving one constraint from one color to the other.  (Fig.~\ref{fig:step-interface} right).

@@

Nielson pointed out  that the two most important issues in designing for usability are understanding the users' tasks and the differences in users \cite{jakob_nielsen_usability_1993}. From analyzing how users use \chameleon{}, we realized that the ideal debugging interface should adapt to the specific programmer and programming task. There are cases where a programmer wants the debugger to simply ``show the answer", and others to dive deeper into the problem domain and search for the optimal solution. To accommodate the need to customize the level of information density and granularity of control, \chameleon{} provides three modes: basic, balanced, and advanced. Programmers can switch between modes by clicking on the mode switching toggles (Fig. \ref{fig:anatomy}-C). The features accessible from different modes are summarised in the following table:%shown in table~\ref{tab:chameleon-features}.

@@

Using only constraints from the MUS is sufficient to locate the type error, but to recover types from type errors we need  constraints from parts of the program that are irrelevant to the type error.  For instance, consider an ill-typed 2-tuple where two possible types can be assigned: \texttt{(Int, Int)} and \texttt{(Int, String)}. The types reconstructed from Chameleon may be \texttt{(a, Int)} and \texttt{(a, String)}. Although the recovered types are theoretically correct, they introduce a new type variable \texttt{a}, making the error message harder to understand. To solve this issue in \chameleon{}, for each constraint \texttt{c} in the MUS, we find a maximally satisfiable subset (MSS) from all the constraints that contain every other element of MUS but not \texttt{c}. These maximally satisfiable subsets, while not helpful to inform error locations, will produce the most concrete conflicting types, see Fig.~\ref{fig:compare-to-original}.

@@

Using concrete types provides extra information to programmers. With a type of \texttt{(Float, Float)}, programmers may want to convey a point in 2d space. However, this information is not preserved in \texttt{(a, Float)}. If the types get too verbose for programmers to identify the conflict between the two signatures, it is always possible to mute the overlapping information and highlight the differences from the debugging front end.

@@

% \begin{lstlisting} [
%     language=Haskell, caption={
%     A simple program that is ill-typed. This program generates two constraints from line 1 and one constraint from line 2.
% }, label={listing:ifelse}
% ]
% if a then b else c
% a = "True"
% \end{lstlisting}
\begin{listing}[!ht]
\begin{minted}{haskell}
    if a then b else c
    a = "True"
\end{minted}
\vspace{-3mm}
\caption{A simple program that is ill-typed. It generates two constraints from line 1 and one constraint from line 2. }
\label{listing:1}
\end{listing}
For instance, for the program in the listing \ref{listing:1}, \chameleon{}  generates the following constraints and labels (in brackets) $T_a = Bool$ (if condition), $T_b = T_c$ (if branches), $T_a= String$  (definition). Clearly, as $T_a$ can not unify with both \textit{Bool} and \textit{String}, this program is not well typed. \chameleon{} can construct a human-readable explanation from the MUS. An example output for listing \ref{listing:1} can be: \texttt{a} has type \texttt{Bool} because \texttt{a} is the condition of an if statement; however, \texttt{a} has type \texttt{String} because \texttt{a} is defined as the string literal \texttt{"True"}. This explanation facilitates the deduction steps (Section \ref{sub:deduction-steps}).

### 2022-12-12 sections/evaluation.tex
Change figure heads

The data collected during study 1a, Fig.~\ref{fig:analysis-1a} does not show significant differences across Tasks 1-7. In hindsight, these tasks were trivial challenges for most users and the individual differences of participants are generally more significant than differences between treatments. However, one interesting observation is task 8, where the \chameleon{} group outperformed the GHC group (p-value = 0.002456, Wilcoxin signed test). We attribute this significant difference to the hardness of Task 8. The source file is longer and involves more language features (abstract data types and high-level functions). GHC struggles to produce a relevant error message for this type of error. 
From this result, we hypothesized that we may observe a more significant difference using tasks with lengthier and more realistic source code. 
This hypothesis is also supported by the most common feedback (7 out of 35) that the tasks were too trivial to invite meaningful evaluation. One participant said, ``Looks nicer than GHC, but without trying it on something more complicated, I cannot conclude whether it would help me in practice."

Therefore, in study 1b we introduced more difficult challenges and indeed observed that the \chameleon{} group were faster (p = 0.00742, Wilcoxin signed test) than the GHC group in almost all tasks (figure \ref{fig:analysis-1b}), barring task 1. For task 1, it is suspected that some participants spent more time exploring the interface of \chameleon{} due to its unfamiliarity. For all other tasks, from the video recordings, we saw many \chameleon{} users confidently skip reading unrelated chunks of code, while GHC users generally read through the whole program. In harder problems and messier code, we notice programmers start to report the benefits of \chameleon{}. ``It's most useful feature that I noticed was that it points out the locations of both conflicting uses; GHC often makes it difficult to figure out how it's coming to a conclusion about a type." reported one participant. ``I think \chameleon{}  does a much better job than GHC's error messages. I like that it shows the sources for the type judgments. This makes it quite easy to figure out how to rectify errors." reported another participant.

@@

\subsubsection{\textbf{\chameleon{} 2}}  \label{sub:us4}
Based on observations of Study 1 we introduced several new features to \chameleon{}, eventually resulting in the UI depicted in Figs.~(2-13). Advanced debugging features such as deduction steps and candidate expressions were available in this iteration. In addition, mode switching is enabled by default. A few other user interfaces were designed and prototyped between the development of \chameleon{} 1 and \chameleon{} 2. Study 2 therefore addresses the research question:

@@

As shown in  Fig.~\ref{fig:r4-analysis}, time to complete each task roughly relates to the interaction level of participants. Participants with higher interaction levels generally performed better, and the lowest interaction level was worse (one-way ANOVA since we have three samples, p = 0.0368). The results from three tasks stand out from the general trend: in Tasks 4 and 6, higher interaction users performed worse, and in task 9, the general trend is exaggerated. As with Study 1a and 1b, this difference is likely related to task difficulty. Tasks 4 and 6 are shorter than other tasks. The ideal fixes for these two tasks are placed relatively early in the source code (both in the first two lines of the source code). These may be useful because following the natural reading order will allow participants to skip lines on the bottom, and thus the advantage of \chameleon{} is less critical. On the other hand, task 9 is the lengthiest task of all. It also involves more complex features, such as mutually recursive type definitions.

@@

Another observation is when using the mode switching feature of \chameleon{}, we find presenting the starting mode and finishing mode of each task and each participant in a confusion matrix is very telling (Fig. \ref{fig:r4-mode-switching}). 
This observation suggests two characteristics of using multi-mode debugging tools. First, programmers are generally reluctant to switch modes (146/277 changing modes vs. 131/277 staying in the same mode). Second, when changing modes, programmers generally switch to the more informative modes instead of, the more concise ones.
\begin{figure}
    \centering
    \includegraphics[width=0.7\linewidth,trim=0mm 5mm 0mm 5mm,clip]{images/mode-switching.pdf}
    \caption{Study 2 mode switches by starting mode.  Users overwhelmingly switched to the more sophisticated interface mode.
        % The rows represent the finishing mode, columns starting modes.
    }
    \label{fig:r4-mode-switching}
\end{figure}

>>>>

The data collected during study 1a, Fig.~\ref{fig:analysis-1a} does not show significant differences across Tasks 1-7. In hindsight, these tasks were trivial challenges for most users and the individual differences among participants are generally more significant than the differences between treatments. However, one interesting observation is task 8, where the \chameleon{} group outperformed the GHC group (p-value = 0.002456, Wilcoxin signed test). We attribute this significant difference to the hardness of Task 8. The source file is longer and involves more language features (abstract data types and high-level functions). GHC struggles to produce a relevant error message for this type of error. 
From this result, we hypothesized that we may observe a more significant difference using tasks with lengthier and more realistic source code. 
This hypothesis is also supported by the most common feedback (7 out of 35) that the tasks were too trivial to invite meaningful evaluation. One participant said, ``Looks nicer than GHC, but without trying it on something more complicated, I cannot conclude whether it would help me in practice."

Therefore, in study 1b we introduced more difficult challenges and indeed observed that the \chameleon{} group was faster (p = 0.00742, Wilcoxin signed test) than the GHC group in almost all tasks (figure \ref{fig:analysis-1b}), barring task 1. For task 1, it is suspected that some participants spent more time exploring the interface of \chameleon{} due to its unfamiliarity. For all other tasks, from the video recordings, we saw many \chameleon{} users confidently skip reading unrelated chunks of code, while GHC users generally read through the whole program. In harder problems and messier code, we notice programmers start to report the benefits of \chameleon{}. ``It's most useful feature that I noticed was that it points out the locations of both conflicting uses; GHC often makes it difficult to figure out how it's coming to a conclusion about a type." reported one participant. ``I think \chameleon{}  does a much better job than GHC's error messages. I like that it shows the sources for the type judgments. This makes it quite easy to figure out how to rectify errors." reported another participant.

@@

\subsubsection{\textbf{\chameleon{} 2}}  \label{sub:us4}
Based on observations of Study 1 we introduced several new features to \chameleon{}, eventually resulting in the UI depicted in Figs.~(2-13). Advanced debugging features such as deduction steps and candidate expressions were available in this iteration. In addition, mode switching is enabled by default. A few other user interfaces \cite{anonymous_interactive_2021} were designed and prototyped between the development of \chameleon{} 1 and \chameleon{} 2. Study 2 therefore addresses the research question:

@@

As shown in  Fig.~\ref{fig:r4-analysis}, the time to complete each task roughly relates to the interaction level of participants. Participants with higher interaction levels generally performed better, and the lowest interaction level was worse (one-way ANOVA since we have three samples, p = 0.0368). The results from three tasks stand out from the general trend: in Tasks 4 and 6, higher interaction users performed worse, and in task 9, the general trend is exaggerated. As with Study 1a and 1b, this difference is likely related to task difficulty. Tasks 4 and 6 are shorter than other tasks. The ideal fixes for these two tasks are placed relatively early in the source code (both in the first two lines of the source code). These may be useful because following the natural reading order will allow participants to skip lines on the bottom, and thus the advantage of \chameleon{} is less critical. On the other hand, task 9 is the lengthiest task of all. It also involves more complex features, such as mutually recursive type definitions.

@@

Another observation is when using the mode switching feature of \chameleon{}, we find presenting the starting mode and finishing mode of each task and each participant in a correlation matrix is very telling (Fig. \ref{fig:r4-mode-switching}). This observation suggests two characteristics of using multi-mode debugging tools. First, programmers are generally reluctant to switch modes (146/277 changing modes vs. 131/277 staying in the same mode). Second, when changing modes, programmers generally switch to the more informative modes instead of, the more concise ones.
\begin{figure}
    \centering
    \includegraphics[width=0.7\linewidth,trim=0mm 5mm 0mm 5mm,clip]{images/mode-switching.pdf}
    \caption{Study 2 mode switches by starting mode.  Users overwhelmingly switched to the more sophisticated interface mode.
        % The rows represent the finishing mode, columns starting modes.
    }
    \label{fig:r4-mode-switching}
\end{figure}

### 2022-12-12 sections/introduction.tex
Updates
Change figure heads

Dynamically-typed programming languages such as JavaScript and Python have risen in popularity in recent decades \cite{chatley_next_2019}. These languages present a low barrier of entry, especially to beginner programmers: they require no type declaration, variable types or object structures can be modified dynamically, and functions can deal with dynamic input using ad-hoc polymorphism and runtime reflection. However, studies show that dynamically-typed languages negatively affect development productivity \cite{kleinschmager_static_2012}, code usability \cite{mayer_empirical_2012}, and code quality \cite{gao_type_2017, ray_large-scale_2017, meyerovich_empirical_2013}. They are often found to produce error-prone code \cite{chen_empirical_2020, wang_empirical_2015,xu_python_2016} and require strong programmer discipline to avoid pitfalls \cite{chen_empirical_2020}. For these reasons, many modern dynamically-typed languages have introduced static typing annotations as part of the core language features in recent years (e.g.\ \textit{TypeScript} and \textit{mypy}).

@@

The challenge of programming in statically-typed languages is that type errors can sometimes be difficult to resolve~\cite{tirronen_understanding_2015}. In particular, they may point to locations that are distant from real error causes, expose errors in hard to comprehend language, or provide misleading fixing suggestions~\cite{wu_how_2017}.
% Compared to dynamic type systems, static type systems offer programmers opportunities to weed out a large number of errors at compile time (reducing the need for run-time debugging) and to increase usability~\cite{mayer_static_2012} and codebase maintainability~\cite{kleinschmager_static_2012}. In recent years, the programming community has seen a trend of migrating away from dynamically typed codebases due to the lack of maintainability and refactoring safety when systems reach large scale and complexity~\cite{chatley_next_2019}. These migration methods include the introduction of gradual type annotations (e.g., JavaScript to TypeScript or Python with mypy) or transition to a modern strongly-typed language (Scala, Rust, Haskell, PureScript, etc.). However, with dynamically typed languages (e.g., JavaScript, PHP, and Python) still being most peoples' formative experience with programming and with such languages now firmly entrenched in education~\cite{stackoverflow_stack_2022}, this transition may pose difficulties for programmers who lack experience writing type-safe programs.

@@

This paper introduces \chameleon{}, an interactive type debugging tool for Haskell. It can visualize the relevant context of a type error (where it happens or could have happened and which parts of code cause it). In addition, \chameleon{} allows programmers to interactively explore all the parts of code where multiple types can be inferred and to resolve ambiguity. The most noticeable features provided by \chameleon{} are the type compare tool, the deduction step feature, and the candidate expression card feature. These features are integrated into a unified debugging environment and can be enabled or disabled separately based on the programmers' preferences and debugging needs. \chameleon{} is open-source and is available in both web~\cite{shuai_fu_chameleon_2022} and desktop versions.  
% \todo{Introduce Haskell earlier?  Motivate WHY Haskell?} The current implementation of \chameleon{} targets the Haskell 2010 language standard. However, it is planned to extend it to support other programming languages using the same underlying ideas and techniques. Historically, Haskell has been a test bed for advanced language features, and it is common for features established in Haskell to be transferred to mainstream programming languages.

@@

This paper makes the following contributions:
\begin{itemize}
\item We provide the design and implementation of the \chameleon{} to visualize the relevant context of a type error and allow programmers to explore and verify the error locations in small chunks interactively.  
\item {
    We report the results of four experiments designed to evaluate \chameleon{}.}
\end{itemize}

>>>>

Dynamically-typed programming languages such as JavaScript and Python have risen in popularity in recent decades \cite{chatley_next_2019}. These languages present a low barrier of entry, especially to beginner programmers: they require no type declaration, variable types or object structures can be modified dynamically, and functions can deal with dynamic input using ad-hoc polymorphism and runtime reflection. However, studies show that dynamically-typed languages negatively affect development productivity \cite{kleinschmager_static_2012}, code usability \cite{mayer_empirical_2012}, and code quality \cite{gao_type_2017, ray_large-scale_2017, meyerovich_empirical_2013}. They are often found to produce error-prone code \cite{chen_empirical_2020, wang_empirical_2015,xu_python_2016} and require strong programmer discipline to avoid pitfalls \cite{chen_empirical_2020}. For these reasons, many modern dynamically-typed languages have introduced static typing annotations as part of the core language features in recent years (e.g.\ \textit{TypeScript}~\cite{microsoft_javascript_nodate} and \textit{mypy}~\cite{mypy_mypy_nodate}).

@@

The challenge of programming in statically-typed languages is that type errors can sometimes be difficult to resolve~\cite{tirronen_understanding_2015, hage_solved_2020}. In particular, they may point to locations that are not the real error causes, expose errors in cryptic language, or provide misleading fixing suggestions~\cite{wu_how_2017}.
% Compared to dynamic type systems, static type systems offer programmers opportunities to weed out a large number of errors at compile time (reducing the need for run-time debugging) and to increase usability~\cite{mayer_static_2012} and codebase maintainability~\cite{kleinschmager_static_2012}. In recent years, the programming community has seen a trend of migrating away from dynamically typed codebases due to the lack of maintainability and refactoring safety when systems reach large scale and complexity~\cite{chatley_next_2019}. These migration methods include the introduction of gradual type annotations (e.g., JavaScript to TypeScript or Python with mypy) or transition to a modern strongly-typed language (Scala, Rust, Haskell, PureScript, etc.). However, with dynamically typed languages (e.g., JavaScript, PHP, and Python) still being most peoples' formative experience with programming and with such languages now firmly entrenched in education~\cite{stackoverflow_stack_2022}, this transition may pose difficulties for programmers who lack experience writing type-safe programs.

@@

This paper introduces \chameleon{}, an interactive type debugging tool for Haskell. It can visualize the relevant context of a type error (where it happens or could have happened and which parts of code cause it). In addition, \chameleon{} allows programmers to interactively explore all the parts of code where multiple types can be inferred and to resolve ambiguity. The most noticeable features provided by \chameleon{} are the type compare tool, the deduction step feature, and the candidate expression card feature. These features are integrated into a unified debugging environment and can be enabled or disabled separately based on the programmers' preferences and debugging needs. \chameleon{} is open-source and is available in both web~\cite{anonymous_chameleon_2022} and desktop versions.  
% \todo{Introduce Haskell earlier?  Motivate WHY Haskell?} The current implementation of \chameleon{} targets the Haskell 2010 language standard. However, it is planned to extend it to support other programming languages using the same underlying ideas and techniques. Historically, Haskell has been a test bed for advanced language features, and it is common for features established in Haskell to be transferred to mainstream programming languages.

@@

This paper makes the following contributions:
\begin{itemize}
\item We provide the design and implementation of the \chameleon{} to visualize the relevant context of a type error and allow programmers to explore and verify the error locations in small chunks interactively.  
\item {
    We report the results of three experiments designed to evaluate \chameleon{}.}
\end{itemize}

### 2022-12-12 sections/related-work.tex
Change figure heads

\subsection{Finding all type error locations}
Many have studied the approach of finding all locations that contribute to a type error (\cite{stuckey_interactive_2003, haack_type_2004, pavlinovic_practical_2015, schilling_constraint-free_2012}). Type error slicing~\cite{haack_type_2004} is a technique that finds locations that are complete and minimal for the type error. Internally labeled constraints and Minimal Unsatisfiable Set (MUS) manipulation are used to generate these slices. The language supported in Haack's work was a subset of Standard ML. The original Chameleon~\cite{stuckey_interactive_2003} used a similar technique but generate constraints in Constraint handling rules (CHR). As a result,  Chameleon supported advanced type-level features (type classes and functionally dependent types). The project also introduced the ability to query type information through a command line-style interface. Although Chameleon was firmly grounded in results from type theory, its designs were never evaluated with user studies. While finding all error locations is useful in comprehending type errors, it is only 1 of the 7 properties listed in the proposed manifesto of good type error reporting~\cite{yang_improved_2000}.

>>>>

\subsection{Finding all type error locations}
Many have studied the approach of finding all locations that contribute to a type error~\cite{stuckey_interactive_2003, haack_type_2004, pavlinovic_practical_2015, schilling_constraint-free_2012}. Type error slicing~\cite{haack_type_2004} is a technique that finds locations that are complete and minimal for the type error. Internally labeled constraints and Minimal Unsatisfiable Set (MUS) manipulation are used to generate these slices. The language supported in Haack's work was a subset of Standard ML. The original Chameleon~\cite{stuckey_interactive_2003} used a similar technique but generate constraints in Constraint handling rules (CHR). As a result,  Chameleon supported advanced type-level features (type classes and functionally dependent types). The project also introduced the ability to query type information through a command line-style interface. Although Chameleon was firmly grounded in results from type theory, its designs were never evaluated with user studies. While finding all error locations is useful in comprehending type errors, it is only 1 of the 7 properties listed in the proposed manifesto of good type error reporting~\cite{yang_improved_2000}.
